---
title: "Technical Vignette to Accompany `Towards reduction in bias in epidemic curves...'"
author: "Burstyn I, Goldstein ND, Gustafson P"
date: "03/04/2020"
output: pdf_document
---

* There are five public files: This .Rmd document, two versions of its pdf output (Alberta and Philadelphia), and two csv files (three-columns: date, positive, negative) for Alberta and Philadelphia. respectively.

* Note the "switch" below to produce either AB or PH output.

* Note the "switch" below to echo the R commands or not in the pdf output.

* Note the "switch" below to additionally produce separate pdf files for manuscript figures. 
 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## which jurisdiction
#JRSDCT <- "AB"
JRSDCT <- "PH"
```

```{r}
SHOW.CODE <- F
### switch to T if want R code echoed in pdf output
```

```{r}
MS.PLOTS <- T
### if true, some plots also output to separate pdf files
```

```{r, echo=SHOW.CODE}
### for nicely formatted pdf figures
if (MS.PLOTS) {
  source("pdfPG.R")  ### author-use only
}
```

```{r, message=F, warning=F}
require(rjags)
require(MCMCvis)
```

```{r, echo=F}
### bespoke function to add CIs to plot
add.ci <- function(x, est, se, lvl=.95, trnc.lo=0) {
  for (i in 1:length(x)) {
    points(rep(x[i],2),
           pmax(est[i]+qnorm(1-(1-lvl)/2)*c(-1,1)*se[i],rep(trnc.lo,2)),
            type="l")
  }
}
```

```{r, echo=SHOW.CODE}
if (JRSDCT=="AB") {
  dta <- read.csv("alberta march 28 brief.csv", header=T)
}
if (JRSDCT=="PH") {
  dta <- read.csv("philadelphia march 31 brief.csv", header=T)
}

T.end <- dim(dta)[1]
  
ystr <- dta$positive[1:T.end]
n <- dta$negative[1:T.end] + dta$positive[1:T.end]
q.hat <- ystr/n
se <- sqrt(q.hat*(1-q.hat)/n)
```

Head and Tail of series

```{r}
head(dta)  # start of series
tail(dta)  # end of series
```

```{r, echo=SHOW.CODE}
### first knot must be at day 1
### last knot must be at day T.end

### choose three-equally spaced interior knots
knts <- c(1, round(T.end*(1:3)/4), T.end)
num.kn <- length(knts)
spc.kn <- knts[-1] - knts[-num.kn]
```

\pagebreak

Proportion positive ($Y^{*}/n$ time-series), with 95% confidence intervals, and knot indicators

```{r, echo=SHOW.CODE}
plot(1:T.end, q.hat, 
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep=""),
     ylab="Proportion Positive", ylim=c(0, max(q.hat+2.1*se)))
add.ci((1:T.end), q.hat, se)
points(knts, rep(0,num.kn),pch=17,col="red")
```

```{r, echo=SHOW.CODE}
if (MS.PLOTS) {
  pdf.PG(paste("figA_",JRSDCT,".pdf",sep=""),1,1, ttlspc=T)
  plot(1:T.end, q.hat, 
       xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep=""),
       ylab="Proportion Positive", ylim=c(0, max(q.hat+2.1*se)))
  add.ci((1:T.end), q.hat, se)
  points(knts, rep(0,num.kn),pch=17,col="red")
  graphics.off()
}  
```

```{r, echo=SHOW.CODE}
### JAGS specification
genmod.JAGS <- "model{
  
  ### prior distribution
  
  ### how much weight on the linear component for sens
  ### (earlier version corresponds to no weight)
  
  ### sn.wt <- 0 no smoothness (as per earlier versions)
  ### sn.wt <- 1 perfectly linear
  sn.wt ~ dunif(0.5,0.9)  ### some unspecified amount of smoothing
  
  ### linear component endpoints
  sn.LL ~ dunif(sn.wt*sn.lo[1], sn.wt*sn.hi[1])
  sn.LR ~ dunif(sn.wt*sn.lo[num.kn], sn.wt*sn.hi[num.kn])
    
  ### prev, sn piecewise linear
  ### parameterized by value at knots
  for (i in 1:num.kn) {
    r.kn[i] ~ dunif(0, r.hi[i])
    
    ### linear component
    sn.L[i] <- ((knts[i]-knts[1])*sn.LR+(knts[num.kn]-knts[i])*sn.LL)/
               (knts[num.kn]-knts[1])
    
    ### jumpy component
    sn.J[i] ~ dunif((1-sn.wt)*sn.lo[i],(1-sn.wt)*sn.hi[i]) 
    
    ### two together
    sn.kn[i] ~ dsum(sn.L[i], sn.J[i])
  }
  
  sp ~ dunif(sp.lo,1)
  
  ### these imply the daily values
  for (i in 1:(num.kn-1)) {
    for (j in 0:(spc.kn[i]-1)) {
      r[knts[i]+j] <- ((spc.kn[i]-j)*r.kn[i]+j*r.kn[i+1])/(spc.kn[i])
    }  
  }    
  r[knts[num.kn]] <- r.kn[num.kn]
  
  ### these imply the daily values
  for (i in 1:(num.kn-1)) {
    for (j in 0:(spc.kn[i]-1)) {
      sn[knts[i]+j] <- ((spc.kn[i]-j)*sn.kn[i]+j*sn.kn[i+1])/(spc.kn[i])
    }  
  }    
  sn[knts[num.kn]] <- sn.kn[num.kn]
  
  for (i in 1:(knts[num.kn])) {
    y[i] ~ dbinom(r[i], n[i])            ### true positives
    ystr1[i] ~ dbinom(sn[i], y[i])       ### correct positives
    ystr2[i] ~ dbinom(1-sp, n[i]-y[i])   ### false positives
    ystr[i] ~ sum(ystr1[i], ystr2[i])
  }

}"
```

```{r message=FALSE, warning=FALSE, results="hide", cache=T, echo=SHOW.CODE}

if (JRSDCT=="AB") {

### specify hyperparameters: 

### important this is inside this (cached) code chunk, 
### so a change in hyperparams triggers MCMC re-run)
### also one such chunk per jurisdiction, for same reason

sn.lo.AB <- rep(0.6, num.kn)  # same bound at each knot
sn.hi.AB <- rep(0.9, num.kn)  # same upper-bound at each knot
r.hi.AB <- rep(0.15, num.kn)   # same upper-bound at each knot
sp.lo.AB <- 0.95

### generative model, data go in

mod <- jags.model(textConnection(genmod.JAGS),
         data=list(knts=knts, num.kn=num.kn, spc.kn=spc.kn,
                   sp.lo=sp.lo.AB, sn.lo=sn.lo.AB, 
                   sn.hi=sn.hi.AB, r.hi=r.hi.AB,
                   ystr=ystr, n=n),
         inits=list(y=round(1.2*ystr), ystr1=ystr, 
                    ystr2=rep(0,length(ystr))),
        n.chains=4)          
                    
###  MC output comes out

opt.JAGS.AB <- coda.samples(mod, 
  variable.names=c("sp","sn.wt","sn.kn","r.kn","y"),
  n.iter=400000, n.thin=200)
}
```


```{r message=FALSE, warning=FALSE, results="hide", cache=T, echo=SHOW.CODE}

if (JRSDCT=="PH") {

### specify hyperparameters: 
### important this is inside this (cached) code chunk, 
### so a change in hyperparams triggers MCMC re-run)
### also one such chunk per jurisdiction, for same reason

sn.lo.PH <- rep(0.6, num.kn)  # same bound at each knot
sn.hi.PH <- rep(0.9, num.kn)  # same upper-bound at each knot
r.hi.PH <- rep(0.5, num.kn)   # same upper-bound at each knot
sp.lo.PH <- 0.95

### generative model, data go in

mod <- jags.model(textConnection(genmod.JAGS),
         data=list(knts=knts, num.kn=num.kn, spc.kn=spc.kn,
                   sp.lo=sp.lo.PH, sn.lo=sn.lo.PH, 
                   sn.hi=sn.hi.PH, r.hi=r.hi.PH,
                   ystr=ystr, n=n),
         inits=list(y=round(1.2*ystr), ystr1=ystr, 
                    ystr2=rep(0,length(ystr))),
        n.chains=4)          
                    
###  MC output comes out

opt.JAGS.PH <- coda.samples(mod, 
  variable.names=c("sp","sn.wt","sn.kn","r.kn","y"),
  n.iter=400000, n.thin=200)
}
```

```{r, echo=SHOW.CODE}
if (JRSDCT=="AB") {
  opt.JAGS <- opt.JAGS.AB
  sn.lo <- sn.lo.AB; sn.hi <- sn.hi.AB
  r.hi <- r.hi.AB; sp.lo <- sp.lo.AB
}  
if (JRSDCT=="PH") {
  opt.JAGS <- opt.JAGS.PH
  sn.lo <- sn.lo.PH; sn.hi <- sn.hi.PH
  r.hi <- r.hi.PH; sp.lo <- sp.lo.PH
}
```

Hyperparameter settings

```{r}
r.hi
sn.lo
sn.hi
sp.lo
```

```{r, echo=SHOW.CODE}
### raw posterior draws for future use
mc.opt <-as.matrix(opt.JAGS)
mc.qnt <- summary(opt.JAGS)$quantiles
n.draws <- dim(mc.opt)[1]
```

\pagebreak

Adjusted daily positive tests

```{r, echo=SHOW.CODE}
ndx <- (1:(dim(mc.qnt)[1]))[row.names(mc.qnt)=="y[1]"]
plot(1:T.end, mc.qnt[ndx:(ndx+T.end-1),"50%"],
     pch=19,
     ylim=c(0, max(mc.qnt[ndx:(ndx+T.end-1),"97.5%"])),
     ylab="Positive Tests",
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep="")
)
for (i in 1:T.end) {
  points(rep(i,2), mc.qnt[ndx+i-1,c("2.5%","97.5%")], type="l")
}

points(1:(knts[num.kn]), ystr, col="blue", pch=5)
```

```{r, echo=SHOW.CODE}
if (MS.PLOTS) {
pdf.PG(paste("figB_",JRSDCT,".pdf",sep=""),1,1, ttlspc=T)  
ndx <- (1:(dim(mc.qnt)[1]))[row.names(mc.qnt)=="y[1]"]
plot(1:T.end, mc.qnt[ndx:(ndx+T.end-1),"50%"],
     pch=19,
     ylim=c(0, max(mc.qnt[ndx:(ndx+T.end-1),"97.5%"])),
     ylab="Positive Tests",
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep="")
)
for (i in 1:T.end) {
  points(rep(i,2), mc.qnt[ndx+i-1,c("2.5%","97.5%")], type="l")
}

points(1:(knts[num.kn]), ystr, col="blue", pch=5)
graphics.off()
}
```

\pagebreak

Specificity in testing pool (posterior distribution)

```{r, eval=F, echo=SHOW.CODE}
mc.qnt["sp",]
```

```{r, echo=SHOW.CODE}
hist(mc.opt[,"sp"], breaks=seq(from=sp.lo,to=1,by=.0025), prob=T,
     xlab="Specificity",ylab="Density",main="") 
```

```{r, echo=SHOW.CODE}
if (MS.PLOTS) {
pdf.PG(paste("figC_",JRSDCT,".pdf",sep=""),1,1, ttlspc=T)  
hist(mc.opt[,"sp"], breaks=seq(from=sp.lo,to=1,by=.0025), prob=T,
     xlab="Specificity",ylab="Density",main="") 
graphics.off()
}
```

\pagebreak

Sensitivity in testing pool (posterior mean and draws)

```{r, echo=SHOW.CODE}
ndx <- sample(1:n.draws, size=75, replace=T)

plot(0,0, type="n", xlim=c(0,T.end), ylim=c(0.5,1),
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep=""), 
     ylab="Sensitivity")

points(knts, sn.lo, type="l", lty=3)
points(knts, sn.hi, type="l", lty=3)

col.start.sn <- (1:(dim(mc.opt)[2]))[colnames(mc.opt)=="sn.kn[1]"]
for (i in ndx) {
  points(knts, mc.opt[i,col.start.sn:(col.start.sn+num.kn-1)],
         type="l", col="grey")
}         

points(knts,
  apply(mc.opt[,col.start.sn:(col.start.sn+num.kn-1)], 2, mean),
  lwd=2, col="red",type="l")
```

```{r}
if (MS.PLOTS) {
pdf.PG(paste("figD_",JRSDCT,".pdf",sep=""),1,1, ttlspc=T)    
plot(0,0, type="n", xlim=c(0,T.end), ylim=c(0.5,1),
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep=""), 
     ylab="Sensitivity")

points(knts, sn.lo, type="l", lty=3)
points(knts, sn.hi, type="l", lty=3)

col.start.sn <- (1:(dim(mc.opt)[2]))[colnames(mc.opt)=="sn.kn[1]"]
for (i in ndx) {
  points(knts, mc.opt[i,col.start.sn:(col.start.sn+num.kn-1)],
         type="l", col="grey")
}         

points(knts,
  apply(mc.opt[,col.start.sn:(col.start.sn+num.kn-1)], 2, mean),
  lwd=2, col="red",type="l")
graphics.off()
}
```

\pagebreak

Prevalence in testing pool (posterior mean and draws)

```{r, echo=SHOW.CODE}
ndx <- sample(1:n.draws, size=50, replace=T)

plot(0,0, type="n", xlim=c(0,T.end), ylim=c(0,max(r.hi)),
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep=""), 
     ylab="Prevalence")
points(knts, r.hi, type="l", lty=3)

col.start.r <- (1:(dim(mc.opt)[2]))[colnames(mc.opt)=="r.kn[1]"]
for (i in ndx) {
  points(knts, mc.opt[i,col.start.r:(col.start.r+num.kn-1)],
         type="l", col="grey")
}         

points(knts,
  apply(mc.opt[,col.start.r:(col.start.r+num.kn-1)], 2, mean),
  lwd=2, col="red",type="l")
```

\pagebreak

$E(Y^{*}/n)=r\times sens + (1-r)*(1-spec)$ (posterior mean and draws)

```{r, echo=SHOW.CODE}
r.kn <- mc.opt[,col.start.r:(col.start.r+num.kn-1)]
r.full <- matrix(NA,dim(mc.opt)[1],T.end)

for (i in 1:(num.kn-1)) {
    for (j in 0:(spc.kn[i]-1)) {
      r.full[,knts[i]+j]<- (1/spc.kn[i])*
        ( (spc.kn[i]-j)*r.kn[,i]+j*r.kn[,i+1] )
    }  
  }    
r.full[,knts[num.kn]] <- r.kn[,num.kn]

sn.kn <- mc.opt[,col.start.sn:(col.start.sn+num.kn-1)]
sn.full <- matrix(NA,dim(mc.opt)[1],T.end)

for (i in 1:(num.kn-1)) {
    for (j in 0:(spc.kn[i]-1)) {
      sn.full[,knts[i]+j]<- (1/spc.kn[i])*
        ( (spc.kn[i]-j)*sn.kn[,i]+j*sn.kn[,i+1] )
    }  
  }    
sn.full[,knts[num.kn]] <- sn.kn[,num.kn]

ndx <- sample(1:n.draws, size=50, replace=T)

plot(0,0, type="n", xlim=c(0,T.end), ylim=c(0,max(q.hat+2.1*se)),
     xlab=paste("Day (",dta$date[1]," - ",dta$date[T.end],")",sep=""), 
     ylab="Expected Count")

for (i in ndx) {
  points(1:T.end, r.full[i,]*sn.full[i,]+ (1-r.full[i,])*(1-mc.opt[i,"sp"]),
         type="l", col="grey")
}         

points(1:T.end,
  apply(r.full*sn.full + (1-r.full)*(1-mc.opt[,"sp"]), 2, mean),
  lwd=2, col="red",type="l")

points(1:T.end, q.hat, ylim=c(0,1))
add.ci((1:T.end), q.hat, se)
```

\pagebreak

Some clues about MCMC numerical performance

```{r}
MCMCtrace(opt.JAGS, params="sp", pdf=F)

MCMCtrace(opt.JAGS, params="sn.wt", pdf=F)

MCMCsummary(opt.JAGS, params="sp")

MCMCsummary(opt.JAGS, params="sn.kn")

MCMCsummary(opt.JAGS, params="r.kn")

MCMCsummary(opt.JAGS, params="y")
```







